---
title: "Aircraft Visualizations by Number of Flights"
author: "Group 5"
title: "Aircraft Visualization"
author: "Group 5"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(ggplot2)
```

### Considerations
1) flights: continuous, carrier delay: continuous.

### Inporting the dataset
```{r import data}
# Import data
aircrafts_ages_raw <- read.csv("ALL_YEARS_DATA.csv")

# filter for aircraft age <= 22
aircrafts_ages_since_2003 <- aircrafts_ages_raw |> 
  filter(AGE <= 22) 

# create flight bins
aircrafts_ages_since_2003 <- aircrafts_ages_since_2003 |>
  mutate(
    flight_bin = cut(
      FLIGHTS,
      breaks = c(0, 100, 500, 1000, 5000, 10000, 25000, 50000, 75000),
      labels = c("0–100", "101–500", "501–1K", "1K–5K", "5K–10K", "10K–25K", "25K–50K", "50K–75K"),
      include.lowest = TRUE
    )
  )
```


### Total Carrier Delay as a Function of Aircraft Total Flights 
How does age by number of flights flown affect total carrier delay?
age: discrete, carrier delay: continuous. 
```{r flights vs tot carrier delay}
ggplot(aircrafts_ages_since_2003, aes(x = flight_bin, y = TOT_CARRIER_DELAY / 60)) +
  geom_point(alpha = 0.3, color = "slateblue") +
  geom_smooth(method = "lm", se = TRUE, color = "pink") +
  labs(
    title = "Total Carrier Delay by Aircraft Age",
    x = "Aircraft Age (number of flights)",
    y = "Total Carrier Delay (hours)"
  ) + 
  scale_x_discrete(  # Fix the x-axis to avoid scientific notation
    labels = function(x) gsub('e[+-][0-9]+', '', x)  # Remove scientific notation
  ) +
  theme(
    plot.title = element_text(hjust = 0.5))
```

### FLIGHTS vs Average Carrier Delay Line Plot
```{r line plot flights vs tot carrier delay}
# Wrangling - Group by flight_bin to calculate average delay
avg_delay_by_flights <- aircrafts_ages_since_2003 |> 
  group_by(flight_bin) |>  # Grouping by flight_bin
  summarize(avg_delay_hrs = mean(TOT_CARRIER_DELAY / 60, na.rm = TRUE))

# Plotting
ggplot(avg_delay_by_flights, aes(x = flight_bin, y = avg_delay_hrs)) + 
  geom_point(color = "plum2", size = 5) + 
  geom_smooth(method = "lm", se = FALSE, color = "springgreen2", linetype = "dashed") + 
  labs(
    title = "Average Carrier Delay by Aircraft Age (in total flights flown)",
    x = "Aircraft Age (number of flights)",
    y = "Average Carrier Delay (hours)"
  ) + 
  scale_y_continuous(
    breaks = c(5, 24, 48, 72, 168, 336),
    labels = c("5 hrs", "1 day", "2 days", "3 days", "1 week", "2 weeks")
  ) + 
  scale_x_discrete(  
    labels = function(x) gsub('e[+-][0-9]+', '', x)  # Removes scientific notation
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

### Transforming the Delay Time
This needs more work
```{r transform delays}
ggplot(aircrafts_ages_since_2003, aes(x = flight_bin, y = log(TOT_CARRIER_DELAY/60))) +
  geom_point(alpha = 0.3, color = "thistle3") +
  geom_smooth(method = "glm", se = TRUE, color = "darkolivegreen3") +
  labs(
    title = "Log-Transformed Carrier Delay by Aircraft Age (as determined by number of flights flown)",
    x = "Aircraft Age (number of flights)",
    y = "log(Total Carrier Delay(hours))"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

```

### BoxPlot for MFR age vs Delays
The delay values vary widely.
```{r blox plot}
ggplot(aircrafts_ages_since_2003, aes(x = factor(flight_bin), y = (TOT_CARRIER_DELAY/60))) +
  geom_boxplot(fill = "thistle3", color = "darkblue", outlier.alpha = 0.2) +
  labs(
    title = "Aircraft Flights vs Carrier Delays",
    x = "Aircraft Age (number of flights)",
    y = "Total Carrier Delay"
  ) +
  scale_y_log10(
    breaks = c(5, 24, 48, 72, 168, 336),
    labels = c("5 hrs", "1 day", "2 days", "3 days", "1 week", "2 weeks")
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Cancellations
```{r wrangling for cancellations}
avg_cancellations_by_flight_bin <- aircrafts_ages_since_2003 |>
  group_by(flight_bin) |>
  summarize(avg_cancellations = mean(COUNT_CARRIER_CANCELLATION, na.rm = TRUE))
```

### Age vs Cancellations Line Plot
```{r cancellations line plot}
# Plotting:
ggplot(avg_cancellations_by_flight_bin, aes(x = flight_bin, y = avg_cancellations)) + 
  geom_point(color = "darkseagreen3", size = 5) +
  geom_smooth(method = "lm", se = FALSE, color = "burlywood3", linetype = "dashed") + #regression line doesn't work
  labs(
    title = "Average Cancellations by Aircraft Age (in total flights flown)",
    x = "Aircraft Age (number of flights)",
    y = "Average Annual Cancellations"
  ) + 
  scale_x_discrete(  # Fix the x-axis to avoid scientific notation
    labels = function(x) gsub('e[+-][0-9]+', '', x)  # Remove scientific notation
  ) + 
  theme(plot.title = element_text(hjust = 0.5))
```
