---
title: "Aircraft Visualization"
author: "Group 5"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(ggplot2)
```


### Considerations
1) airtime: continuous, carrier delay: continuous.
2) all the plots are thrown off by the NA values

### Importing the dataset

```{r import data}
# Import data
aircrafts_ages_raw <- read.csv("ALL_YEARS_DATA.csv")

# filter for aircraft age <= 22
aircrafts_ages_since_2003 <- aircrafts_ages_raw |> 
  filter(AGE <= 22) 

# create flight bins
aircrafts_ages_since_2003 <- aircrafts_ages_since_2003 |>
  mutate(
    airtime_bin = cut(
      AIR_TIME,
      breaks = c(0, 100, 500, 1000, 5000, 10000, 25000, 50000, 75000),
      labels = c("0–100", "101–500", "501–1K", "1K–5K", "5K–10K", "10K–25K", "25K–50K", "50K–75K")
    )
  )

# Wrangling
avg_delay_by_airtime_age <- aircrafts_ages_since_2003 |>
  group_by(airtime_bin) |>
  summarize(avg_delay_hrs = mean(TOT_CARRIER_DELAY / 60, na.rm = TRUE))
```


### Total Carrier Delay as a Function of Aircraft Airtime 
How does age as determined by time spent in the air affect total carrier delay?
```{r airtime, age tot carrier delay}
ggplot(aircrafts_ages_since_2003, aes(x = airtime_bin, y = TOT_CARRIER_DELAY / 60)) +
  geom_point(alpha = 0.3, color = "slateblue", na.rm = TRUE) +
  geom_smooth(method = "lm", se = TRUE, color = "pink", na.rm = TRUE) +
  labs(
    title = "Total Carrier Delay by Total Hours Flown by an Aircraft",
    x = "Aircraft Age (airtime)",
    y = "Total Carrier Delay (hours)"
  ) + 
  scale_x_discrete(  # Fix the x-axis to avoid scientific notation
    labels = function(x) gsub('e[+-][0-9]+', '', x)  # Remove scientific notation
  ) + scale_y_continuous(
    breaks = c(5, 24, 48, 72, 168),
    labels = c("5 hrs", "1 day", "2 days", "3 days", "1 week")
  ) +
  
  theme(
    plot.title = element_text(hjust = 0.5))

#TODO: THE NA VALUES 
```

### AIRTIME vs Average Carrier Delay Line Plot
```{r line plot airtime vs tot carrier delay}

ggplot(avg_delay_by_airtime_age, aes(x = airtime_bin, y = avg_delay_hrs)) +
  geom_point(color = "plum2", size = 5) +
  geom_smooth(method = "lm", se = FALSE, color = "springgreen2", linetype = "dashed") +
  labs(
    title = "Average Carrier Delay by Aircraft Age",
    x = "Aircraft Age (time spent flying)",
    y = "Average Carrier Delay (hours)"
  ) +
  scale_y_continuous(
    breaks = c(5, 24, 48, 72, 168, 336),
    labels = c("5 hrs", "1 day", "2 days", "3 days", "1 week", "2 weeks")
  ) +
  theme(plot.title = element_text(hjust = 0.5))

```

### Transforming the Delay Time
```{r transform delays}
ggplot(aircrafts_ages_since_2003, aes(x = airtime_bin, y = log(TOT_CARRIER_DELAY / 60))) +
  geom_point(alpha = 0.3, color = "thistle3") +
  geom_smooth(method = "glm", se = TRUE, color = "darkolivegreen3") +
  labs(
    title = "Log-Transformed Carrier Delay by Aircraft Age",
    x = "Aircraft Airtime Bin (hours)",
    y = "log(Total Carrier Delay (hours))"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```

### BoxPlot for MFR age vs Delays
The delay values vary widely.
```{r blox plot}
ggplot(aircrafts_ages_since_2003, aes(x = airtime_bin, y = (TOT_CARRIER_DELAY/60))) +
  geom_boxplot(fill = "thistle3", color = "darkblue", outlier.alpha = 0.2) +
  labs(
    title = "Aircraft Total Airtime vs Carrier Delays",
    x = "Aircraft Age (in time spent flying)",
    y = "Total Carrier Delay"
  ) +
  scale_y_continuous(
    breaks = c(5, 24, 48, 72, 168, 336),
    labels = c("5 hrs", "1 day", "2 days", "3 days", "1 week", "2 weeks")
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Cancellations
```{r wrangling for cancellations}
# Wrangling
avg_cancellations_by_age <- aircrafts_ages_since_2003 |>
  group_by(airtime_bin) |>
  summarize(avg_cancellations = mean(COUNT_CARRIER_CANCELLATION, na.rm = TRUE))
```

### Age vs Cancellations Line Plot
```{r cancellations line plot}
ggplot(avg_cancellations_by_age, aes(x = airtime_bin, y = avg_cancellations)) +
  geom_line(color = "burlywood3") +
  geom_point(color = "darkseagreen3", size = 5) +
  labs(
    title = "Average Cancellations by Aircraft Total Airtime",
    x = "Aircraft Age (in time spent flying)",
    y = "Average Annual Cancellations"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```



