---
title: "Logistic Regression"
author: "Group 5"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, messages=FALSE}
library(tidyverse)
library(openintro)
library(ggplot2)
library(dplyr)
```

### Import data
Ratios_2 <- read.csv("Ratios_2.csv")



### Scatterplot which shows the observed proportion of planes that were delayed over 30 mins vs Aircraft FLIGHTS
```{r prop planes delayed over 30 mins}
Ratios_2$FLIGHTS_BIN <- round(Ratios_2$FLIGHTS, -3)  # e.g., round to nearest 1000

FLIGHTS_means <- Ratios_2 |>
  group_by(FLIGHTS_BIN) |>
  summarize(
    COUNT_CARRIER_DELAY_30_RATIO = mean(COUNT_CARRIER_DELAY_30_RATIO, na.rm = TRUE)
  )

real_data_plot <- ggplot(FLIGHTS_means, aes(x=FLIGHTS_BIN, y=COUNT_CARRIER_DELAY_30_RATIO)) +
  geom_point() +
  labs(x = "FLIGHTS of Aircraft (total)", y = "Proportion Delayed 30 mins")

plot(real_data_plot)
```


### Cubic Model (best)
```{r cubic model}
# Remove top 1% of FLIGHTS values
cutoff <- quantile(Ratios_2$FLIGHTS, 0.99, na.rm = TRUE)
Ratios_trimmed <- Ratios_2 |> filter(FLIGHTS <= cutoff)


# Step 1: Fit the cubic model on the raw data
m3 <- glm(COUNT_CARRIER_DELAY_30_RATIO ~ FLIGHTS + I(FLIGHTS^2) + I(FLIGHTS^3),
          data = Ratios_trimmed, family = binomial)

# Step 2: Create the plot of observed (binned) data
plot(FLIGHTS_means$FLIGHTS_BIN, FLIGHTS_means$COUNT_CARRIER_DELAY_30_RATIO,
     xlab = "FLIGHTS of Aircraft (binned to nearest 1000)",
     ylab = "Proportion Delayed 30 mins",
     pch = 16, col = "blue",
     main = "Observed Data with Cubic Model")

# Step 3: Create a smooth sequence of FLIGHTS for predictions
FLIGHTS_seq <- seq(min(FLIGHTS_means$FLIGHTS_BIN), max(FLIGHTS_means$FLIGHTS_BIN), length.out = 200)

# Step 4: Predict using cubic model coefficients
coefs3 <- coef(m3)
logits3 <- coefs3["(Intercept)"] + 
           coefs3["FLIGHTS"] * FLIGHTS_seq + 
           coefs3["I(FLIGHTS^2)"] * FLIGHTS_seq^2 + 
           coefs3["I(FLIGHTS^3)"] * FLIGHTS_seq^3
probs3 <- exp(logits3) / (1 + exp(logits3))

# Step 5: Overlay the model prediction line
lines(FLIGHTS_seq, probs3, col = "red", lwd = 2)

```


*****


### Linear model for FLIGHTS vs Cancellations
```{r cubic model cancellations}
m1.5 <- glm(Proportion_Cancelled ~ FLIGHTS, data = Ratios_2, family = binomial)
summary(m1.5)
```

### Linear model plot for FLIGHTS vs cancellations 
The real values do seem to follow a linear pattern.
These predicted probabilities are very small (between 0.013% and 0.13%).
The prediction curve is basically flat.
The model sees almost no meaningful relationship between aircraft FLIGHTS and cancellation rate
``` {r cubic model for cancellations plot}
# Plot the observed data
plot(FLIGHTS_means$FLIGHTS, FLIGHTS_means$Proportion_Cancelled,
     xlab = "FLIGHTS of Aircraft (years)",
     ylab = "Proportion Cancelled",
     pch = 16, col = "blue",
     main = "Observed Data with Linear Regression Model")

# Sequence of FLIGHTSs for smooth prediction curves
FLIGHTS_seq <- seq(min(FLIGHTS_means$FLIGHTS), max(FLIGHTS_means$FLIGHTS), length.out = 200)

# Linear model predictions
coefs1.5 <- coef(m1.5)
logits1.5 <- coefs1.5["(Intercept)"] + coefs1.5["FLIGHTS"] * FLIGHTS_seq
probs1.5 <- exp(logits1.5) / (1 + exp(logits1.5))

# Add model prediction line for m1
lines(FLIGHTS_seq, probs1.5, col = "red", lwd = 2)

# Add horizontal dashed line at mean observed proportion delayed
abline(h = mean(FLIGHTS_means$Proportion_Cancelled), col = "gray", lty = 2)

# printing the probability to explain why the line is flat
range(probs1.5)
```
