---
title: "Logistic Regression"
author: "Group 5"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(ggplot2)
library(dplyr)
```

### Import data
Ratios_2 <- read.csv("Ratios_2.csv")



<!-- ### Scatterplot which shows the observed proportion of planes that were delayed over 30 mins vs Aircraft AGE -->
<!-- ```{r prop planes delayed over 30 mins} -->
<!-- age_means <- aggregate(COUNT_CARRIER_DELAY_30_RATIO ~ AGE, data = Ratios_2, FUN = mean) -->

<!-- real_data_plot <- ggplot(age_means, aes(x=AGE, y=COUNT_CARRIER_DELAY_30_RATIO)) + -->
<!--   geom_point() + -->
<!--   labs(x = "Age of Aircraft (years)", y = "Proportion Delayed 30 mins") -->

<!-- plot(real_data_plot) -->
<!-- ``` -->

<!-- ### Linear Regression Model using AGE as the only predictor -->
<!-- ```{r math} -->
<!-- m1 <- glm(COUNT_CARRIER_DELAY_30_RATIO ~ AGE, data = Ratios_2, family = binomial) -->
<!-- summary(m1) -->
<!-- ``` -->

<!-- ### PLotting Linear Regression Model -->
<!-- ```{r linear model plotting} -->
<!-- # Plot the observed data -->
<!-- plot(age_means$AGE, age_means$COUNT_CARRIER_DELAY_30_RATIO, -->
<!--      xlab = "Age of Aircraft (years)", -->
<!--      ylab = "Proportion Delayed 30 mins", -->
<!--      pch = 16, col = "blue", -->
<!--      main = "Observed Data with Linear Regression Model") -->

<!-- # Sequence of ages for smooth prediction curves -->
<!-- age_seq <- seq(min(age_means$AGE), max(age_means$AGE), length.out = 200) -->

<!-- # Linear model predictions -->
<!-- coefs1 <- coef(m1) -->
<!-- logits1 <- coefs1["(Intercept)"] + coefs1["AGE"] * age_seq -->
<!-- probs1 <- exp(logits1) / (1 + exp(logits1)) -->

<!-- # Add model prediction line for m1 -->
<!-- lines(age_seq, probs1, col = "red", lwd = 2) -->

<!-- # Add horizontal dashed line at mean observed proportion delayed -->
<!-- abline(h = mean(age_means$COUNT_CARRIER_DELAY_30_RATIO), col = "gray", lty = 2) -->
<!-- ``` -->

<!-- ### Quadratic Model using AGE and AGE^2 -->
<!-- and testing it: a 5 year old plane should have a avg delay proportion of 0.0389 -->
<!-- ```{r quadratic model} -->
<!-- # quadratic model using both mage and mageˆ2 as predictors -->
<!-- m2 <- glm(COUNT_CARRIER_DELAY_30_RATIO ~ AGE + I(AGE^2), data = Ratios_2, family = binomial) -->
<!-- summary(m2) -->

<!-- # Calculate logit and probability for a 5-year-old plane -->
<!-- m2logit_5 <- coef(m2)["(Intercept)"] + coef(m2)["AGE"] * 5 + coef(m2)["I(AGE^2)"] * 5^2 -->
<!-- print(m2logit_5) -->
<!-- # Convert logit to probability -->
<!-- m2prob_5 <- exp(m2logit_5) / (1 + exp(m2logit_5)) -->
<!-- print(m2prob_5) -->
<!-- ``` -->

<!-- ### Plotting Quadratic Model -->
<!-- ```{r quadratic model plotting} -->
<!-- # Plot the observed data -->
<!-- plot(age_means$AGE, age_means$COUNT_CARRIER_DELAY_30_RATIO, -->
<!--      xlab = "Age of Aircraft (years)", -->
<!--      ylab = "Proportion Delayed 30 mins", -->
<!--      pch = 16, col = "blue", -->
<!--      main = "Observed Data with Logistic Regression Model") -->

<!-- # Sequence of ages for smooth prediction curves -->
<!-- age_seq <- seq(min(age_means$AGE), max(age_means$AGE), length.out = 200) -->

<!-- # Quadratic model predictions -->
<!-- coefs2 <- coef(m2) -->
<!-- logits2 <- coefs2["(Intercept)"] + coefs2["AGE"] * age_seq + coefs2["I(AGE^2)"] * age_seq^2 -->
<!-- probs2 <- exp(logits2) / (1 + exp(logits2)) -->

<!-- # Add model prediction line for m2 -->
<!-- lines(age_seq, probs2, col = "darkgreen", lwd = 2) -->

<!-- # Add horizontal dashed line at mean observed proportion delayed -->
<!-- abline(h = mean(age_means$COUNT_CARRIER_DELAY_30_RATIO), col = "gray", lty = 2) -->
<!-- ``` -->

### Cubic Model (best)
```{r cubic model}
m3 <- glm(COUNT_CARRIER_DELAY_30_RATIO ~ AGE + I(AGE^2) + I(AGE^3), 
          data = Ratios_2, 
          family = binomial)
summary(m3)
```
### Plotting cubic model Age vs Delay30
```{r cubic model plot}
# real data 
plot(age_means$AGE, age_means$COUNT_CARRIER_DELAY_30_RATIO,
     xlab = "Age of Aircraft (years)",
     ylab = "Proportion Delayed 30 mins",
     pch = 16, col = "blue",
     main = "Observed Data with Cubic Model")

# Sequence of ages for smooth prediction curves
age_seq <- seq(min(age_means$AGE), max(age_means$AGE), length.out = 200)

# Cubic model predictions
coefs3 <- coef(m3)
logits3 <- coefs3["(Intercept)"] + 
           coefs3["AGE"] * age_seq + 
           coefs3["I(AGE^2)"] * age_seq^2 + 
           coefs3["I(AGE^3)"] * age_seq^3
probs3 <- exp(logits3) / (1 + exp(logits3))


lines(age_seq, probs3, col = "red", lwd = 2)
```

<!-- ### Quartic Model -->
<!-- ```{r quartic model} -->
<!-- m4 <- glm(COUNT_CARRIER_DELAY_30_RATIO ~ AGE + I(AGE^2) + I(AGE^3) + I(AGE^4), -->
<!--           data = Ratios_2, family = binomial) -->
<!-- summary(m4) -->
<!-- ``` -->

<!-- ### Plotting Quartic Model -->
<!-- there is no significant difference between the quartic model graph and the cubic model graph -->
<!-- ```{r quartic model plot} -->
<!-- # Plot the observed data -->
<!-- plot(age_means$AGE, age_means$COUNT_CARRIER_DELAY_30_RATIO, -->
<!--      xlab = "Age of Aircraft (years)", -->
<!--      ylab = "Proportion Delayed 30 mins", -->
<!--      pch = 16, col = "blue", -->
<!--      main = "Observed Data with Quartic Model") -->

<!-- # Sequence of ages for smooth prediction curves -->
<!-- age_seq <- seq(min(age_means$AGE), max(age_means$AGE), length.out = 200) -->

<!-- # Quartic model predictions -->
<!-- coefs4 <- coef(m4) -->
<!-- logits4 <- coefs4["(Intercept)"] +  -->
<!--            coefs4["AGE"] * age_seq +  -->
<!--            coefs4["I(AGE^2)"] * age_seq^2 +  -->
<!--            coefs4["I(AGE^3)"] * age_seq^3 +  -->
<!--            coefs4["I(AGE^4)"] * age_seq^4 -->
<!-- probs4 <- exp(logits4) / (1 + exp(logits4)) -->

<!-- # Add model prediction line -->
<!-- lines(age_seq, probs4, col = "darkgreen", lwd = 2) -->
<!-- ``` -->


*****


### Linear model for Age vs Cancellations
```{r cubic model cancellations}
m1.5 <- glm(Proportion_Cancelled ~ AGE, data = Ratios_2, family = binomial)
summary(m1.5)
```

### Linear model plot for age vs cancellations 
The real values do seem to follow a linear pattern.
These predicted probabilities are very small (between 0.013% and 0.13%).
The prediction curve is basically flat.
The model sees almost no meaningful relationship between aircraft age and cancellation rate
``` {r cubic model for cancellations plot}
# Plot the observed data
plot(age_means$AGE, age_means$Proportion_Cancelled,
     xlab = "Age of Aircraft (years)",
     ylab = "Proportion Cancelled",
     pch = 16, col = "blue",
     main = "Observed Data with Linear Regression Model")

# Sequence of ages for smooth prediction curves
age_seq <- seq(min(age_means$AGE), max(age_means$AGE), length.out = 200)

# Linear model predictions
coefs1.5 <- coef(m1.5)
logits1.5 <- coefs1.5["(Intercept)"] + coefs1.5["AGE"] * age_seq
probs1.5 <- exp(logits1.5) / (1 + exp(logits1.5))

# Add model prediction line for m1
lines(age_seq, probs1.5, col = "red", lwd = 2)

# Add horizontal dashed line at mean observed proportion delayed
abline(h = mean(age_means$Proportion_Cancelled), col = "gray", lty = 2)

# printing the probability to explain why the line is flat
range(probs1.5)
```
